A highly-available move operation for replicated trees

### 2021-03-02 paper/review-response.tex

## How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic?
The paper is well structured. I particularly liked section 2 that clearly illustrates the main issues with concurrent non-coordinated move operations. Also, by showing that the replicated filesystems of Dropbox and Google drive both have bugs demonstrates how difficult is to devise a correct resolution policy for the move operation. Overall, section 2 nicely shows the work relevance and potential impact without getting too technical.
The replication algorithm is thoroughly described and the proof of correctness is presented at the right level of detail - although the proofs are an essential component a more detailed presentation would become tedious.

## Please rate the readability of the manuscript.
The paper is well written and easy to follow.

## Detailed comments

C1: The system model and network requirements should be explicitly defined.
\end{spverbatim}
\authorcomment{We have improved the description of the system model at the beginning of Section~\ref{sec:algorithm}.}
\begin{spverbatim}
C2: Is the metadata parameter necessary? It seems that this parameter could be abstracted away which would simplify the algorithm presentation.
\end{spverbatim}
\authorcomment{The metadata is not needed for the functioning of our algorithm per se, but it is needed for several applications of our algorithm, such as filesystems (metadata is a filename) or JSON/XML (metadata indicates the order of children of the same parent).
Since it does not significantly complicate the algorithm, we would prefer to keep it as-is.}
\begin{spverbatim}
C3: Further justification is needed for choosing to encode add and remove operations as move operations. My intuition, base on previously published results as CISE and Hamsaz, is that not all operations need to follow a total ordered across replicas. However, because all operations are defined as moves, conflicts are always possible. I was left wondering if by having explicit add and remove operations some undo-do-redo cycles could be avoided.
\end{spverbatim}
\authorcomment{We have done substantial further work on this question, and added Section~\ref{sec:create} to discuss it.
In particular, we have proved in Isabelle that creation operations can skip the undo/redo process, provided that some additional assumptions are met.
We now also explain why the same optimisation does not work for remove operations.}
\begin{spverbatim}
C4: Instead of having an undo-do-redo cycle would it be possible to have state checkpoints? As operations might arrive in any order, the number of operations to undo can be high, in that case, it might be useful to have state checkpoints and combine these with undo operations.
\end{spverbatim}
\authorcomment{Periodically checkpointing the state would indeed reduce the number of operations to undo, since a replica could jump to the checkpoint nearest to the log position where a new operation needs to be inserted.
However, we believe that this would not bring about a large performance improvement, because we would still have to perform the redo step for all operations with timestamps greater than the new operation, which still has a cost that is proportional to the number of operations we have to undo if there are no checkpoints.

@@

This redo is necessary because the new operation may cause any subsequent operation to flip from safe to unsafe or vice versa, and so we have to recompute that safety status for all operations with timestamps greater than the new operation.}
\begin{spverbatim}
C5: Notation is not consistent. It is stated that types are primed letters (2nd paragraph of section 3.1), however, this naming convention is dropped later on.
\end{spverbatim}
\authorcomment{We have carefully checked the paper and confirmed that the notation is correct.

Primed letters are \emph{type variables}, not concrete types.
For example, the type of our tree is written in Isabelle as $('n \times '\!m \times '\!n)\;\mathit{set}$, which means a set of triples where the first and third element of each triple have the same type (since they both represent a tree node, namely the parent and the child). 
The type variable $'n$ can then be instantiated with any concrete type; for example, in our evaluation it is a 64-bit integer.}
\begin{spverbatim}
C6: The evaluation proves the point that the tree CRDT is the best option for fast response time and disconnected mode (in this case it is the only option).

### 2021-03-02 paper/move-op.tex
Revision submitted to TPDS

\IEEEPARstart{M}{any} applications use a tree-structured data model.
For example, most file systems are trees: directories are branch nodes, and files are leaf nodes.
Moreover, XML and JSON documents are trees, and they are used in many applications to represent e.g.\ rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
Typically, a graphical user interface allows a user to edit this information interactively, resulting in updates to the underlying XML/JSON structure: adding or deleting nodes in the tree, and moving nodes from one position in the tree to another.

@@

The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, and Nokia Bell Labs.
Our evaluation was conducted using AWS credits from the AWS Educate program.
Thank you to Marc Shapiro for feedback on a draft of this paper.

>>>>

\IEEEPARstart{M}{any} applications use a tree-structured data model.
Most filesystems are trees: directories are branch nodes, files are leaf nodes (we discuss hardlinks in \S\ref{sec:extensions}).
XML and JSON documents are also trees, and they are used in many applications to represent e.g.\ rich text (a tree of paragraphs, lists, figures, sections, etc.), vector graphics, CAD drawings, and many other types of data.
Typically, a graphical user interface allows a user to edit this information interactively, resulting in updates to the underlying XML/JSON structure: adding or deleting nodes in the tree, and moving nodes from one position in the tree to another.

@@

The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship, the Isaac Newton Trust, Nokia Bell Labs, and crowdfunding supporters including Molham Aref, Chet Corcos, Mintter.com, David Pollak, SoftwareMill, Talent Formation Network, and Adam Wiggins.
Our evaluation was conducted using AWS credits from the AWS Educate program.
Thank you to Marc Shapiro for feedback on a draft of this paper.
